\hspace{1cm} Do syntactic and semantic similarity lead to interference effects? \newline Evidence from self-paced reading and event-related potentials using German

### 2023-12-27 paper/main.tex
minor edits

It is well-established in sentence processing that completing linguistic dependencies requires memory retrieval. As an example, consider 
\ref{ex:1_vandyke07}.

@@

Here, we consider two kinds of cues used in retrieval: semantic and  syntactic cues.
In a sentence like \ref{ex:1_vandyke07}, the retrieval cues generated at the verb are standardly assumed to include at least  the semantic cue \{+animate\} and the syntactic cue \{+grammatical subject\}. These cues pick out the target noun \textit{the resident} as the subject. The distractor \textit{warehouse}, intervening between subject and verb, does not match either of these retrieval cues. However, if the distractor were \textit{neighbor} instead of \textit{warehouse}, then it would match the \{+animate\} cue at the verb. This overlap in cues between the target noun and the distractor is called cue-overload in cue-based retrieval accounts, and is predicted to trigger semantic interference, which expresses itself as greater processing time at the verb compared to the baseline condition \ref{ex:1_vandyke07}.

By contrast, syntactic interference is expected to be triggered if the distractor matches the \{+grammatical subject\} cue. This situation can occur when the relative clause  in \ref{ex:1_vandyke07} that modifies \textit{the resident} were to be changed to \textit{who said that the warehouse was dangerous}. Now, the distractor matches the \{+grammatical subject\} cue.

@@

Previous work has shown clear evidence for this kind of retrieval interference \parencite{vandyke07, vandyke_mcelree06, vandyke_mcelree2011, nicenboim,  vandyke_lewis03,jaeger_etal_2017}.
However, two important questions remain unanswered. First, as discussed in \textcite{jaeger_etal_2017}, almost all the published studies on interference effects are severely underpowered. Low power has the consequence that published estimates will tend to be overestimates or misestimates of the effect size \parencite{vasishth2018_signficancefilter}. Over-/misestimates have important implications for theory evaluation. One example is discussed in \textcite{vasishth2018_signficancefilter}, where a key prediction of surprisal theory \parencite{levy&keller_2013} could not be validated when a larger-sample study was conducted. By contrast, the original small-sample study \parencite{levy&keller_2013} showed suggestive evidence consistent with the theory. Another example is the recent heated debate on prediction in sentence comprehension, which was largely focused on the results of low-powered studies, leading to potentially misleading conclusions \parencite{nicenboim_etal_2020}.
Second, although it has been established \parencite[e.g.,][]{vandyke07,mertzen} that syntactic and semantic cues are involved in sentence processing, it is not obvious which cues are relevant for a particular retrieval event. For example, \textcite{dillon2013} argued, following \textcite{Sturt2003}, that the gender cue was either downweighted or not used at all when an antecedent-reflexive dependency needs to be built. Instead, the antecedent of a reflexive could be identified using the syntactic principle A of the binding theory, which correctly identifies the target for retrieval.  Using computational modeling, \textcite{dillon2013} suggested that antecedent-reflexive dependencies were immune to interference, and that the reason for this is that principle A of the binding theory is the primary driver of the cue-based retrieval process \parencite[cf. ][]{jaeger_etal_2020,yadav2021individual}. This was a surprising discovery because a priori, cue-based retrieval accounts do not allow for exceptions to retrieval interference in particular syntactic dependency types.
More broadly, an important implication of the work initiated by \textcite{dillon2013}, among others, is that  the precise details of which cues are used during retrieval can affect processing difficulty in possibly surprising ways.

@@

As the same materials were used for all our experiments, we describe them here once before we describe the specifics of the experiments, separately. The studies had 120 items with four conditions in a Latin square design shown in \ref{ex:materials}. The items were constructed following \citeauthor{vandyke07}'s (\citeyear{vandyke07}) items and partially re-used  \citeauthor{mertzen}'s (\citeyear{mertzen}) items.

@@

The comprehension of the sentences in \ref{ex:materials} requires a memory retrieval operation at the verb \textit{trank} (`drank') to retrieve its non-adjacent subject \textit{Witwer} (`widower'). During this retrieval, the distractor \textit{Einbrecher} (`burglar') / \textit{Verlust} (`loss') might cause interference and impede comprehension.

Syntactic interference is manipulated via the syntactic status of the distractor, which is either subject (\ref{ex:hisyn}) of an embedded clause, or part of a prepositional phrase (\ref{ex:losyn}). Following \textcite{vandyke07}, the assumption here is that  during the retrieval of the subject, a distractor in any subject position causes more syntactic interference than a distractor inside a prepositional phrase.

@@

\paragraph{Materials in the plausibility study}
For each of the 120 item quadruplets, four simple sentences with subject-verb-object word order were constructed, so that each of the nouns in the item once functioned as the subject of the critical verb. The norming materials corresponding to the item depicted in \ref{ex:materials} are shown in \ref{ex:materials_norming}.

@@

Self-paced reading times were analyzed with Bayesian linear mixed models in R \citep{r} with log-normal likelihood, using the brms package \citep{brms}. The critical word for analyses was the verb \textit{trank} `drank' in \ref{ex:materials}, constituting the retrieval site. Because previous work found effects in the pre-critical and post-critical region, reading times of the pre-critical word (\textit{regelmäßig} `regularly' in \ref{ex:materials}) and the spill-over region (\textit{um} `in.order.to' in \ref{ex:materials}) were analyzed. 
The models included fixed effects for syntactic interference (high +0.5, low -0.5), semantic interference (high +0.5, low -0.5) and their interaction.  Varying intercepts for participants and items were included; varying slopes were excluded for reasons explained below.

@@

In order to quantify the uncertainty on the parameters of interest, we report 95\% credible intervals.  
These intervals represent the range over which we can be 95\% certain that the values of the parameter lies, given the statistical model and the data. However, formal hypothesis testing cannot be carried out without a likelihood ratio test that compares two alternative models \parencite{schad_etal_2022_BF,Royall}. For this reason, formal hypothesis tests for the presence or absence of the effects of interest were carried out using Bayes factors. Because Bayes factors can be very sensitive to prior specifications on the target parameter being tested \parencite{schad_etal_2022_BF}, 
we report a sensitivity analysis using a range of prior specifications for the relevant parameters (see Table \ref{tab:spr_priors}). The priors assume a priori effect sizes for the main effects of syntactic and semantic interference, and their interaction, ranging from 0-8 ms, 0-40 ms, or 0-81 ms.  These priors are based on the observed sign and sizes of effects and uncertainties in previous reading studies that use the present design \parencite{vandyke07,mertzen}.

The Bayes factor is analogous to the frequentist ANOVA or likelihood ratio test: it compares the (marginal) likelihood of a full model to that of a baseline null model; this is often written as $BF_{10}$, with the subscripts indicating that marginal likelihood of the full model is in the numerator and that of the null model is in the denominator. One major advantage of the Bayes factor over the frequentist analogue is that it takes the uncertainty of the parameters into account. This leads to more conservative inferences compared to frequentist ANOVA, which only takes the maximum likelihood estimate of the parameter into account \parencite[see][for detailed discussion]{schad_etal_2022_BF}. Another important advantage of the Bayes factor -- one that is very relevant for the present work -- is that it is in principle possible to find evidence both for or against an effect. This stands in contrast to the frequentist ANOVA, which is designed to only furnish evidence against the null.

>>>>

It is well-established in sentence processing that completing linguistic dependencies requires memory retrieval. As an example, consider 
(\ref{ex:1_vandyke07}).

@@

Here, we consider two kinds of cues used in retrieval: semantic and  syntactic cues.
In a sentence like (\ref{ex:1_vandyke07}), the retrieval cues generated at the verb are standardly assumed \parencite[e.g.,][]{vandyke07} to include at least  the semantic cue \{+animate\} and the syntactic cue \{+grammatical subject\}. These cues pick out the target noun \textit{the resident} as the subject. The distractor \textit{warehouse}, intervening between subject and verb, does not match either of these retrieval cues. However, if the distractor were \textit{neighbor} instead of \textit{warehouse}, then it would match the \{+animate\} cue at the verb. This overlap in cues between the target noun and the distractor is called cue-overload in cue-based retrieval accounts, and is predicted to trigger semantic interference, which expresses itself as greater processing time at the verb compared to the baseline condition (\ref{ex:1_vandyke07}).

By contrast, syntactic interference is expected to be triggered if the distractor matches the \{+grammatical subject\} cue. This situation can occur when the relative clause  in (\ref{ex:1_vandyke07}) that modifies \textit{the resident} were to be changed to \textit{who said that the warehouse was dangerous}. Now, the distractor matches the \{+grammatical subject\} cue.

@@

Previous work has shown clear evidence for this kind of retrieval interference \parencite{vandyke07, vandyke_mcelree06, vandyke_mcelree2011, nicenboim,  vandyke_lewis03,jaeger_etal_2017}.
However, two important questions remain unanswered. First, as discussed in \textcite{jaeger_etal_2017}, almost all the published studies on interference effects are severely underpowered. Low power has the consequence that published estimates will tend to be overestimates or misestimates of the effect size \parencite{vasishth2018_signficancefilter}. Over-/misestimates have important implications for theory evaluation. One example is discussed in \textcite{vasishth2018_signficancefilter}, where a key prediction of surprisal theory \parencite{levy&keller_2013} could not be validated when a larger-sample study was conducted. By contrast, the original small-sample study \parencite{levy&keller_2013} showed suggestive evidence consistent with the theory. Another example is the recent heated debate \parencite{nieuwland_etal_2018} on prediction in sentence comprehension, which was largely focused on the results of low-powered studies, leading to potentially misleading conclusions \parencite{nicenboim_etal_2020}.
Second, although it has been established \parencite[e.g.,][]{vandyke07,mertzen} that syntactic and semantic cues are involved in sentence processing, it is not obvious which cues are relevant for a particular retrieval event. For example, \textcite{dillon2013} argued, following \textcite{Sturt2003}, that the gender cue was either downweighted or not used at all when an antecedent-reflexive dependency needs to be built. Instead, the antecedent of a reflexive could be identified using the syntactic principle A of the binding theory, which correctly identifies the target for retrieval.  Using computational modeling, \textcite{dillon2013} suggested that antecedent-reflexive dependencies were immune to interference, and that the reason for this is that principle A of the binding theory is the primary driver of the cue-based retrieval process \parencite[cf. ][]{jaeger_etal_2020,yadav2021individual}. This was a surprising discovery because the default assumption in cue-based retrieval accounts was that all cues were equally weighted, regardless of the dependency type; this had the consequence that retrieval interference was expected whenever a cue-overload configuration was present in a sentence. More broadly, an important insight in the work of \textcite{dillon2013}, among others, is that  the precise details of which cues are used during retrieval can affect processing difficulty in possibly surprising ways. We build on this idea in the present paper.

@@

As the same materials were used for all our experiments, we describe them here once before we describe the specifics of the experiments, separately. The studies had 120 items with four conditions in a Latin square design shown in (\ref{ex:materials}). The items were constructed following \citeauthor{vandyke07}'s (\citeyear{vandyke07}) items and partially re-used  \citeauthor{mertzen}'s (\citeyear{mertzen}) items.

@@

The comprehension of the sentences in (\ref{ex:materials}) requires a memory retrieval operation at the verb \textit{trank} (`drank') to retrieve its non-adjacent subject \textit{Witwer} (`widower'). During this retrieval, the distractor \textit{Einbrecher} (`burglar') / \textit{Verlust} (`loss') might cause interference and impede comprehension.

Syntactic interference is manipulated via the syntactic status of the distractor, which is either the subject (\ref{ex:hisyn}) of an embedded clause, or part of a prepositional phrase (\ref{ex:losyn}) and hence a non-subject. Following \textcite{vandyke07}, the assumption here is that  during the retrieval of the subject, a distractor in \emph{any} subject position causes more syntactic interference than a distractor inside a prepositional phrase.

@@

\paragraph{Materials in the plausibility study}
For each of the 120 item quadruplets, four simple sentences with subject-verb-object word order were constructed, so that each of the nouns in the item once functioned as the subject of the critical verb. The norming materials corresponding to the item depicted in (\ref{ex:materials}) are shown in (\ref{ex:materials_norming}).

@@

Self-paced reading times were analyzed with Bayesian linear mixed models in R \citep{r} with log-normal likelihood, using the brms package \citep{brms}. The critical word for analyses was the verb \textit{trank} `drank' in (\ref{ex:materials}), constituting the retrieval site. Because previous work found effects in the pre-critical and post-critical region, reading times of the pre-critical word (\textit{regelmäßig} `regularly' in \ref{ex:materials}) and the spill-over region (\textit{um} `in.order.to' in \ref{ex:materials}) were analyzed. 
The models included fixed effects for syntactic interference (high +0.5, low -0.5), semantic interference (high +0.5, low -0.5) and their interaction.  Varying intercepts for participants and items were included; varying slopes were excluded for reasons explained below.

@@

In order to quantify the uncertainty on the parameters of interest, we report 95\% credible intervals.  
These intervals represent the range over which we can be 95\% certain that the values of the parameter lies, given the statistical model and the data. However, formal hypothesis testing cannot be carried out without a likelihood ratio test that compares two alternative models \parencite{schad_etal_2022_BF,Royall}. For this reason, formal hypothesis tests for the presence or absence of the effects of interest were carried out using Bayes factors. Because Bayes factors can be very sensitive to prior specifications on the target parameter being tested \parencite{schad_etal_2022_BF}, 
we report a sensitivity analysis using a range of prior specifications for the relevant parameters (see Table \ref{tab:spr_priors}). The priors assume a priori effect sizes for the main effects of syntactic and semantic interference, and their interaction, ranging from 0-8 ms, 0-40 ms, or 0-81 ms.  These priors are based on the observed sign and sizes of effects and uncertainties in previous reading studies that use the present design \parencite{vandyke07,mertzen}, and on meta-analyses relating to interference effects  \parencite{jaeger_etal_2017}. For the syntactic and semantic interference effects, we use directional priors so that any evidence for the effect of interest from Bayes factors can be interpreted as evidence of replication success \parencite{verhagen2014bayesian}.

The Bayes factor is analogous to the frequentist ANOVA or likelihood ratio test: it compares the marginal likelihood of a full model to that of a baseline null model; this is often written as $BF_{10}$, with the subscripts indicating that marginal likelihood of the full model is in the numerator and that of the null model is in the denominator. One major advantage of the Bayes factor over the frequentist analogue is that it takes the uncertainty of the parameters into account. This leads to more conservative inferences compared to frequentist ANOVA, which only takes the maximum likelihood estimate of the parameter into account \parencite[see][for detailed discussion]{schad_etal_2022_BF}. Another important advantage of the Bayes factor -- one that is very relevant for the present work -- is that it is in principle possible to find evidence both for or against an effect. This stands in contrast to the frequentist ANOVA which, in its standard usage, is designed to only furnish evidence against the null.
